Abstractions for Type-Level Programming

### 2022-04-22 ack.tex
Remove diff

I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for agreeing to review my dissertation and for taking part in the committee.

>>>>

I would also like to thank the other members of my thesis committee, Paolo Giarrusso, Richard Eisenberg, and Viktor Kuncak, for reviewing my dissertation and for providing valuable feedback.

### 2022-04-22 article.tex
Remove diff

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
\begin{diff}
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representatives for abstractions and type abstractions, respectively.
\end{diff}
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
\begin{diff}
This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$, with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
Although $\Γ \⊢ \S \⇌ \T$ witnesses mutual subtyping between $\S$ and $\T$, that relation is asymmetric, since its left-hand side is always a match type.
\end{diff}
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
\begin{diff}
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} (disjointness/subtyping exclusivity) in the presence of empty types.
\end{diff}

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
\begin{diff}
There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.
\end{diff}

@@

\begin{diff}
In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotations of higher-kinded types.
In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.

@@

At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
However, it would be conceivable to lift this restriction in the future, as shown by the analysis presented in this section.
\end{diff}

@@

\begin{diff}
Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
Erasure hence creates an expressivity gap between pattern matching at the term and at the type level.

In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning that the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.
\end{diff}

@@

\begin{diff}
In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.
\end{diff}

@@

\begin{diff}
For example, consider the following function definition:
%
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
%
In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int

def foo[X](x: X): Foo[X] = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
\end{diff}

>>>>

\subsubsection*{Disjointness / Subtyping Exclusivity}
The following non-standard lemma is necessary to prevent overlap between the \SMatch1/2 and \SMatch3/4 rules.
%
\input{proofs/disjointness-subtyping-exclusivity.tex}
%
If such overlap would be allowed, match types could reduce in several different ways, resulting in an unsound system.
We prove \Cref{lem:disjointness-subtyping-exclusivity} by contradiction.
Our proof uses a mapping from \SystemFm's types into non-empty subsets of a newly defined set $\P \= \{ \Λ, \V \} \∪ \C$, where $\Λ$ and $\V$ are equivalence class representatives for abstractions and type abstractions, respectively.
Elements of $\P$ can be understood as equivalence classes for \Fm's types.
We show that the subtyping relation in \Fm corresponds to a subset relation in $\P$, and that the type disjointness relation in \Fm ($\disj$) corresponds to set disjointness in $\P$.
This set-theoretical view lets us conclude the desired result directly.
In our Coq mechanization, we axiomatize this lemma and delegate to the pen-and-paper proof.

@@

The following \Cref{lem:inversion-of-subtyping} allows us to perform inversion on the subtyping relation,
which is important to show canonical forms (\Cref{lem:canonical-forms}) and inversion of typing (\Cref{lem:inversion-of-typing}).
Stating the lemma requires the definition of a new relation denoted $\Γ \⊢ \S \⇌ \T$, defined in \Cref{fig:harpoon}.
This relation represents evidence of the mutual subtyping between a match type $\S$ and a type $\T$, with the additional constraint that this evidence was exclusively constructed using pairwise applications of \SMatch1/2, \SMatch3/4, and \STrans in both directions.
Intuitively, $\Γ \⊢ \S \⇌ \T$ is handier than two independent derivations of $\Γ \⊢ \S \<: \T$ and $\Γ \⊢ \T \<: \S$ because it allows simultaneous induction on both subtyping directions.
Although $\Γ \⊢ \S \⇌ \T$ witnesses mutual subtyping between $\S$ and $\T$, that relation is asymmetric, since its left-hand side is always a match type.
%
\input{proofs/inversion-of-subtyping.tex}
%
The first point of \Cref{lem:inversion-of-subtyping} uses the structure of the $\⇌$ to provide a form of inversion, which we use to prove each of the subsequent points.
In comparison with the corresponding inversion lemma in \Fsub's safety proof, the statement and the proof of \Cref{lem:inversion-of-subtyping} are longer and more intricate.
This difference is inevitable, given that match type reduction rules allow match expressions to be typed as the result of their reduction, which complexifies the inversion.

@@

An important limitation of \SystemFm compared to Scala's type system is that it does not support empty types.
The bottom of Scala's subtyping lattice, called |Nothing|, provides a direct way to refer to empty sets of values.
Intersection types also provide a way to construct uninhabited types given that Scala does not forbid intersecting two disjoint types.
Empty types are problematic for the match type reduction algorithm as they break the fundamental assumption that two types cannot be both disjoint and subtypes (\Cref{lem:disjointness-subtyping-exclusivity}).
To account for this, our implementation uses an additional \emph{inhabitance check} on scrutinee types before attempting to prove types disjoint.
This updated definition of disjointness allows us to maintain the property of \Cref{lem:disjointness-subtyping-exclusivity} (disjointness/subtyping exclusivity) in the presence of empty types.

@@

Scala tuples, for example, fall into this category.
|Tuple2|, the class for pairs, is defined as follows:
%
\begin{lstlisting}
case class Tuple2[+T1, +T2](_1: T1, _2: T2)
\end{lstlisting}
%
Given two instantiations of this this type, |Tuple2[Int,X]| and |Tuple2[String,X]|, although |Tuple2[Nothing,X]| is a subtype of both, there is no runtime value of type |Tuple2[Nothing,X]| (since |Nothing| is uninhabited), and, as a result, those two types are disjoint.
Our disjointness algorithm implements this kind of reasoning to conclude disjointness in the presence of covariant type parameters.
There is no such reasoning available for contravariant types: our algorithm can never conclude that |G[X]| and |G[Y]| are disjoint for a contravariant type |G|.

@@

In \Cref{subsec:disjointness-of-variant-types} we discussed how our match type reduction algorithm handles variance annotations of higher-kinded types.
In this section, we discuss the orthogonal problem of supporting variance annotations on match type definitions.

@@

At the time of writing, our implementation does not support variance annotations on match types (all match types are invariant).
However, it would be conceivable to lift this restriction in the future, as shown by the analysis presented in this section.

@@

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
Erasure hence creates an expressivity gap between pattern matching at the term and at the type level.

In practice, the Scala compiler handles this limitation by emitting warnings for unchecked patterns.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning that the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.

@@

In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.

@@

For example, consider the following function definition:
%
\begin{lstlisting}
def foo[X](x: X) = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}
%
In the absence of a result type annotation, Scala's type inference will first assign \lstinline!String|Int! for the match expression, and then immediately widen that union to the least upper bound of left- and right-hand-side, |Any|, which will then be used as |foo|'s (inferred) result type.
To obtain a precise result type for this method, users should annotate it explicitly:
%
\begin{lstlisting}
type Foo[X] = X match
  case Int => String
  case String => Int

def foo[X](x: X): Foo[X] = x match
  case x: Int => x.toString
  case x: String => x.toInt
\end{lstlisting}

### 2022-04-22 preprint.tex
Remove diff

\begin{diff}
A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
Although the ideas presented in this chapter did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala language.
\end{diff}

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
\begin{diff}
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
\footnotetext{%
When a generalized singleton type contains a component of form \texttt{(\_: T)}, that type is not singleton: it might contains more than one value.}
\end{diff}

@@

Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
\begin{diff}
However, in the case of a |dependent| method, an explicit result type only serves as an upper bound used to type-check the method's body.
\end{diff}
At other call sites, the (precise) inferred result type is used.
Bounds of dependent methods are written using a special syntax (|<: T|).

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type.
However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
\begin{diff}
More generally, when a dependently-typed function calls into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.
\end{diff}

@@

\begin{diff}
Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
\citeauthor{cardelli1988structural}'s type system is much more expressive than ours and allows bounded quantification over both types and terms, using the notion of a |Type| type and power types.
Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.

@@

Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
Though they introduce a form of dependency, path-dependent types were not designed for type-level computations, rendering their goals largely orthogonal to ours.
\end{diff}

>>>>

A prototype of our Scala extension is available online, in a branch of the dotty-staging repository\footnote{\texttt{git clone git@github.com:dotty-staging/dotty.git --branch add-transparent-7}}.
Although the ideas presented in this chapter did not reach the production stage, our prototype influenced the design of match types (\Cref{chap:match-types}) which are now part of the Scala language.

@@

When inferring the result type of a |dependent| method, our system lifts the method's body into a generalized singleton type.
This lifting will be precise for the subset of expressions that is representable in types, and approximative for the rest.
When we encounter an unsupported construct, we compute its type using the default mode, yielding a type |T| which we then integrate in the lifted body as |_: T|\footnotemark.
\footnotetext{%
When a generalized singleton type contains a component of form \texttt{(\_: T)}, that type is not singleton: it might contains more than one value.}

@@

Scala requires recursive methods to have an explicit result type, and this restriction also applies to |dependent| methods.
However, in the case of a |dependent| method, an explicit result type only serves as an upper bound used to type-check the method's body.
At other call sites, the (precise) inferred result type is used.
Bounds of dependent methods are written using a special syntax (|<: T|).

@@

Scala's type system permits uncontrolled side effects in programs.
Given the absence of an effect system, result types of methods do not convey any information about the potential use of side effects in the method body.
The situation is analogous for |dependent| methods.
Since we uniformly approximate all side effects, we avoid the situation where a type refers to a value that may be modified during the program execution.
For instance, if |z| is a mutable integer variable, we will never introduce |z| in a singleton type.
However, we can still assign singleton types to expressions containing |z|, for example, we can type |Cons(z, Nil())| as |{ Cons(_: Int, Nil()) }|.
More generally, when a dependently-typed function calls into a non-dependently-typed one, we approximate the type of that call to |(_: T)|, where |T| is the  declared result type of the non-dependently-typed function.

@@

Dependently-typed lambda calculi with subtyping were described at least as far back as 1988~\citep{cardelli1988structural}.
\citeauthor{cardelli1988structural}'s type system is much more expressive than ours and allows bounded quantification over both types and terms, using the notion of a |Type| type and power types.
Unlike our system, which is designed with the concrete evaluation of types in mind, Cardelli does not provide a semantics for his system and leaves the equivalence relation among types unspecified.

@@

Dependent-object types~\citep{amin2017type} model the core of Scala's type system and include type members and path-dependent types, which are not represented in our formalism.
Though they introduce a form of dependency, path-dependent types were not designed for type-level computations, rendering their goals largely orthogonal to ours.

### 2022-04-22 thesis.tex
Remove diff

In Scala~3, the rules for implicit ambiguities changed to take implicit parameters into account, thus introducing another, more direct, disambiguation mechanism.
All else being equal, an implicit definition that takes some implicit parameters is considered to be less specific than an implicit definition that takes none~\citep[Changes in Implicit Resolution]{odersky2013scala}.
As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).

@@

% TODO: Compare to singleton types

@@

% TODO you could write a bit more. Maybe move some of the text of the conclusions of individual chapters there?

>>>>

In Scala~3, the rules for implicit ambiguities changed to take implicit parameters into account, thus introducing another, more direct, disambiguation mechanism.
All else being equal, an implicit definition that takes implicit parameters is considered less specific than an implicit definition does not take implicit parameters~\citep[Changes in Implicit Resolution]{odersky2013scala}.
As a result, implicit-based type-level programs written in Scala~3 do not need to use the patterns of ambiguities and priorities presented in this section (\ref{subsec:implicit-priorities} and \ref{subsec:implicit-ambiguities}).

@@

@@
